---
title: "Post processing and visualising MAST results - all thresholds"
output: html_document
date: "2025-03-25"
author: "Sofie Agerbæk"
---

```{r libraries, include = FALSE}
library(Biostrings)
library(ape)
library(seqinr)
library(tidyverse)
library(dplyr)
library(TreeDist)
library(MetBrewer)
library(phytools)
library(patchwork)
library(zoo)
```

```{r theme_setup, include = FALSE}
library(cowplot)
options(digits = 3)
options(pillar.sigfig = 3)

text_base_size <- 12
fig.witdh      <- 180  # mm
fig.height     <- 125  # mm

theme_set(theme_cowplot(font_size = 12, rel_small = 1, rel_tiny = 1, rel_large = 1))
knitr::opts_chunk$set(fig.width  = fig.witdh  / 25.4)
knitr::opts_chunk$set(fig.height = fig.height / 25.4)
knitr::opts_chunk$set(dpi = 600)

ggplot_text_size <- text_base_size / ggplot2::.pt
```

```{r palettes, include = FALSE}
pall_global <- met.brewer("Klimt", n = 6)
options(ggplot2.discrete.fill   = pall_global)
options(ggplot2.discrete.colour = pall_global)
options(ggplot2.continuous.colour = scales::seq_gradient_pal(low = "white", high = "#023F62"),
        ggplot2.continuous.fill   = scales::seq_gradient_pal(low = "white", high = "#023F62"))
```


## Threshold configuration

One entry per threshold. To add or change a threshold, edit this chunk only — everything
downstream reads from `threshold_configs`.

- `threshold`:   the gap threshold string (also used in input/output filenames)
- `output_dir`:  folder name inside `output/`
- `best_k`:      optimal number of trees by BIC minimum (verified against BIC CSV below)
- `top_trees`:   tree numbers in the best model, in descending weight order
                 (from the iterative MAST run in `iqtree_modelselect.R`)

Trees in descending weight order per threshold (all 35):
0.5 → 30 8 31 10 2 21 6 9 4 24 14 29 16 26 19 32 20 13 1 23 25 7 12 35 28 33 27 5 17 11 18 22 3 34 15
0.9 → 5 33 25 17 24 9 22 2 15 4 32 35 10 ...
0.1 → 28 22 27 13 9 8 16 23 7 26 ...

```{r threshold_config}
threshold_configs <- list(
  list(
    threshold  = "0.1",
    output_dir = "0.1",
    best_k     = 4,           # BIC minimum at n=4 (178240.07)
    top_trees  = c(28, 22, 27, 13)
  ),
  list(
    threshold  = "0.5",
    output_dir = "0.5_output",
    best_k     = 6,           # BIC minimum at n=6 (173132.59)
    top_trees  = c(30, 8, 31, 10, 2, 21)
  ),
  list(
    threshold  = "0.9",
    output_dir = "0.9",
    best_k     = 5,           # BIC minimum at n=5 (163767.40)
    top_trees  = c(5, 33, 25, 17, 24)
  )
)
```

Figures are written to `../figures/<threshold>/`, mirroring the `output/` and `generated/`
layout. Create those subdirectories up front:
```{r figures_dirs}
for (cfg in threshold_configs) {
  dir.create(paste0("../figures/", cfg$threshold), showWarnings = FALSE, recursive = TRUE)
}
```

Every figure below is saved as PNG (or PDF, for the already-vector ones) plus a matching
SVG. These three helpers wrap the save calls so each plot only needs one line:
```{r save_helpers}
library(svglite)

# svglite (like most R vector devices) declares physical size in the SVG's
# width/height attributes using "pt" (1/72in). Several common SVG viewers/editors
# (older Inkscape releases especially) misread that "pt" suffix and treat the
# number as unitless px at 96dpi instead of 72dpi, silently shrinking the
# displayed size to 72/96 = 75% of the true physical size. Rewriting those two
# attributes to explicit, unambiguous "mm" (same physical size, just relabelled)
# avoids the whole class of bug. Called on every SVG we write, right after it's
# written.
set_svg_mm <- function(path, width_mm, height_mm) {
  txt <- readLines(path, warn = FALSE)
  txt <- sub(
    "width='[0-9.]+pt' height='[0-9.]+pt'",
    sprintf("width='%.2fmm' height='%.2fmm'", width_mm, height_mm),
    txt
  )
  writeLines(txt, path)
}

mm_of <- function(value, units) {
  value * switch(units, "mm" = 1, "cm" = 10, "in" = 25.4, stop("unknown units: ", units))
}

# ggplot/patchwork object -> raster PNG + vector SVG at the same physical size.
save_gg <- function(plot, path_no_ext, width, height, units = "mm", dpi = 400, bg = "white") {
  ggsave(paste0(path_no_ext, ".png"), plot = plot, device = "png",
         width = width, height = height, units = units, dpi = dpi, bg = bg)
  ggsave(paste0(path_no_ext, ".svg"), plot = plot, device = "svg",
         width = width, height = height, units = units, bg = bg)
  set_svg_mm(paste0(path_no_ext, ".svg"), mm_of(width, units), mm_of(height, units))
}

# draw_fn(): a zero-arg function that draws onto whatever device is currently open
# (print(some_ggplot), or base plot() calls e.g. for cophylo objects) -> PDF + SVG.
save_pdf_and_svg <- function(draw_fn, path_no_ext, width, height, ...) {
  pdf(paste0(path_no_ext, ".pdf"), width = width, height = height, ...)
  draw_fn()
  dev.off()
  svglite(paste0(path_no_ext, ".svg"), width = width, height = height)
  draw_fn()
  dev.off()
  set_svg_mm(paste0(path_no_ext, ".svg"), mm_of(width, "in"), mm_of(height, "in"))
}

# Same, but PNG instead of PDF (for base-graphics plots previously saved as PNG only).
save_raster_and_svg <- function(draw_fn, path_no_ext, width, height, units = "in", res = 300, bg = "white") {
  scale <- switch(units, "in" = 1, "mm" = 1 / 25.4, "cm" = 1 / 2.54, stop("unknown units: ", units))
  w_in  <- width  * scale
  h_in  <- height * scale
  png(paste0(path_no_ext, ".png"), width = w_in, height = h_in, units = "in", res = res, bg = bg)
  draw_fn()
  dev.off()
  svglite(paste0(path_no_ext, ".svg"), width = w_in, height = h_in, bg = bg)
  draw_fn()
  dev.off()
  set_svg_mm(paste0(path_no_ext, ".svg"), mm_of(w_in, "in"), mm_of(h_in, "in"))
}
```

Each of the 35 candidate trees is built from a fixed-position genome window (18 non-overlapping
windows, plus 17 more shifted by half a window — see `scripts/make_ML_trees.Rmd`). This reads
those window boundaries back off the `aln_segment_<start>_to_<end>.treefile` filenames, in the
same order they were originally concatenated into `35_trimmed_ML_trees.newick` (18-set first,
then the 17-set), so row `tree_num == i` matches tree `i` everywhere else in this script:
```{r window_helper}
get_tree_windows <- function(threshold) {
  parse_dir <- function(dir) {
    files  <- unique(basename(list.files(
      dir, pattern = "^aln_segment_[0-9]+_to_[0-9]+\\.treefile$", recursive = TRUE
    )))
    m      <- regmatches(files, regexec("^aln_segment_([0-9]+)_to_([0-9]+)\\.treefile$", files))
    starts <- as.integer(vapply(m, `[[`, character(1), 2))
    ends   <- as.integer(vapply(m, `[[`, character(1), 3))
    data.frame(start = starts, end = ends)[order(starts), ]
  }
  windows <- rbind(
    parse_dir(paste0("../generated/", threshold, "/18_ML_trees")),
    parse_dir(paste0("../generated/", threshold, "/17_ML_trees"))
  )
  windows$tree_num <- seq_len(nrow(windows))
  windows
}
```

Sanity check — verify that the hardcoded `best_k` matches the BIC minimum in each CSV:
```{r bic_sanity_check}
for (cfg in threshold_configs) {
  bic <- read.csv(paste0("../output/", cfg$output_dir, "/bic_values.csv"), header = TRUE)
  auto_k <- bic$n[which.min(bic$value)]
  if (auto_k != cfg$best_k) {
    warning(sprintf(
      "Threshold %s: hardcoded best_k = %d but BIC minimum is at n = %d. Check the config!",
      cfg$threshold, cfg$best_k, auto_k
    ))
  } else {
    message(sprintf("Threshold %s: best_k = %d confirmed by BIC data.", cfg$threshold, cfg$best_k))
  }
}
```


---

## Per-threshold analyses

### BIC vs. number of trees

```{r bic_plots, fig.show='hold'}
for (cfg in threshold_configs) {

  BIC_table <- read.csv(
    paste0("../output/", cfg$output_dir, "/bic_values.csv"),
    header = TRUE
  ) %>% mutate(BIC_transformed = value / 10000)

  bic_plot <- ggplot(BIC_table, aes(x = n, y = BIC_transformed)) +
    geom_point(colour = "black", size = 1.2) +
    geom_vline(xintercept = cfg$best_k, linetype = "dashed", colour = "grey50") +
    theme_bw() +
    labs(title = paste0("Gap threshold: ", cfg$threshold, "  |  best k = ", cfg$best_k)) +
    theme(
      text                = element_text(size = 8),
      axis.ticks          = element_blank(),
      panel.grid.minor    = element_blank(),
      axis.text           = element_text(size = 8, colour = "black"),
      axis.title          = element_blank(),
      plot.title          = element_text(size = 8)
    )

  print(bic_plot)

  save_gg(bic_plot, paste0("../figures/", cfg$threshold, "/BIC_plot"),
          width = 65, height = 50, units = "mm", dpi = 400, bg = "white")
}
```

Same BIC curve, with each power-bootstrap replicate's own BIC-vs-k curve drawn faintly
behind it (see `power_bootstrap/`: B = 100 simulated replicates per threshold, refit over
k = 2..10). If the empirical minimum sits where the replicate curves also bottom out, the
selected k is a stable feature of the data rather than a coin flip:
```{r bic_power_overlay, fig.show='hold'}
for (cfg in threshold_configs) {

  BIC_table <- read.csv(
    paste0("../output/", cfg$output_dir, "/bic_values.csv"),
    header = TRUE
  ) %>% mutate(BIC_transformed = value / 10000)

  rep_bic <- read.csv(paste0(
    "../power_bootstrap/results/", cfg$threshold, "/bic_by_rep_", cfg$threshold, ".csv"
  )) %>% mutate(BIC_transformed = BIC / 10000)

  bic_power_plot <- ggplot(BIC_table, aes(x = n, y = BIC_transformed)) +
    geom_line(
      data = rep_bic, aes(x = k, y = BIC_transformed, group = rep),
      colour = "grey70", alpha = 0.15, linewidth = 0.3, inherit.aes = FALSE
    ) +
    geom_line(colour = "black", linewidth = 0.4) +
    geom_point(colour = "black", size = 1.2) +
    geom_vline(xintercept = cfg$best_k, linetype = "dashed", colour = "grey50") +
    theme_bw() +
    labs(title = paste0("Gap threshold: ", cfg$threshold, "  |  best k = ", cfg$best_k)) +
    theme(
      text                = element_text(size = 8),
      axis.ticks          = element_blank(),
      panel.grid.minor    = element_blank(),
      axis.text           = element_text(size = 8, colour = "black"),
      axis.title          = element_blank(),
      plot.title          = element_text(size = 8)
    )

  print(bic_power_plot)

  save_gg(bic_power_plot, paste0("../figures/", cfg$threshold, "/BIC_plot_with_power"),
          width = 65, height = 50, units = "mm", dpi = 400, bg = "white")

  message("Done: BIC + power-bootstrap overlay for threshold ", cfg$threshold)
}
```


### Gappiness and entropy of alignments

```{r gappy_entropy_plots, fig.show='hold'}

# Store per-threshold data for re-use in the site likelihood section below
gapdf_list     <- list()
entropydf_list <- list()

site_entropy_fn <- function(M, gaps = c("-", "N", "n"), base = 2) {
  apply(M, 2, function(col) {
    col <- col[!(col %in% gaps)]
    if (!length(col)) return(NA_real_)
    p <- prop.table(table(col))
    -sum(p * log(p, base = base))
  })
}

for (cfg in threshold_configs) {

  aln_path <- paste0("../data/", cfg$threshold, "_trimmed_alns_HIV_2021_subset.fasta")
  window   <- 20

  # --- Gappiness ---
  aln     <- readDNAMultipleAlignment(aln_path)
  aln_mat <- as.matrix(aln)
  gaps    <- colSums(aln_mat == "-") / nrow(aln_mat)

  gapdf           <- data.frame(gap_prop = gaps)
  gapdf$site      <- as.numeric(row.names(gapdf))
  gapdf$rollingmean <- zoo::rollapply(gapdf$gap_prop, width = window, mean, partial = TRUE)

  # --- Entropy ---
  dna       <- read.dna(aln_path, format = "fasta")
  H_nogap   <- site_entropy_fn(as.character(dna))
  entropydf <- data.frame(H_nogap = H_nogap)
  entropydf$site        <- as.numeric(rownames(entropydf))
  entropydf$rollingmean <- zoo::rollapply(entropydf$H_nogap, width = window, mean, fill = NA)

  # Store for site likelihood section
  gapdf_list[[cfg$threshold]]     <- gapdf
  entropydf_list[[cfg$threshold]] <- entropydf

  # --- Plots ---
  gapprop_plot <- ggplot(gapdf) +
    geom_point(aes(x = site, y = gap_prop), size = 0.7, color = "grey80") +
    geom_line( aes(x = site, y = rollingmean), color = "grey50", alpha = 0.9) +
    scale_x_continuous(expand = c(0,0),
                       breaks = c(0,1000,2000,3000,4000,5000,6000,7000,8000,9000)) +
    scale_y_continuous(breaks = c(0, 0.25, 0.5)) +
    theme(
      text         = element_text(size = 8),
      axis.title   = element_blank(),
      axis.ticks.y = element_blank(),
      axis.line.x  = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.x  = element_blank()
    )

  entropy_plot <- ggplot(entropydf, aes(x = site)) +
    geom_point(aes(y = H_nogap), size = 0.7, alpha = 0.4, color = "grey80") +
    geom_line( aes(y = rollingmean), color = "grey50", alpha = 0.9) +
    scale_x_continuous(expand = c(0,0),
                       breaks = c(0,1000,2000,3000,4000,5000,6000,7000,8000,9000)) +
    scale_y_continuous(breaks = c(0, 1, 2)) +
    theme(
      text             = element_text(size = 8),
      axis.title       = element_blank(),
      axis.ticks.y     = element_blank(),
      legend.position  = 'bottom'
    ) +
    guides(colour = 'none')

  comb_plot <- gapprop_plot / entropy_plot
  print(comb_plot)

  save_gg(comb_plot, paste0("../figures/", cfg$threshold, "/gappy_entropy"),
          width = 190, height = 70, units = "mm", dpi = 300, bg = "white")
}
```


### Tree distances (MDS of best-model trees)

```{r tree_distance_plots, fig.show='hold'}
for (cfg in threshold_configs) {

  k         <- cfg$best_k
  top_trees <- cfg$top_trees
  pall_k    <- met.brewer("Klimt", n = k)

  trees <- read.tree(
    paste0("../output/", cfg$output_dir,
           "/mast_model_", k, "/mast_model_final_", k, ".treefile")
  )

  pair_dist <- MutualClusteringInfo(trees)
  mds_trees <- cmdscale(as.dist(pair_dist))

  mds_df <- data.frame(
    x       = mds_trees[, 1],
    y       = mds_trees[, 2],
    tree_nr = as.character(top_trees)
  )

  tree_dist_plot <- ggplot(mds_df, aes(x = x, y = y, colour = tree_nr)) +
    geom_point(size = 3) +
    labs(
      title  = paste0("Tree distances — threshold: ", cfg$threshold),
      colour = NULL, y = NULL, x = NULL
    ) +
    scale_colour_manual(values = pall_k) +
    theme(
      text        = element_text(size = 8),
      axis.ticks  = element_blank(),
      plot.title  = element_text(size = 8)
    )

  print(tree_dist_plot)

  save_gg(tree_dist_plot, paste0("../figures/", cfg$threshold, "/tree_distances"),
          width = 85, height = 55, units = "mm", dpi = 400, bg = "white")
}
```


### Site likelihood — modal classes

For each threshold, reads the best-k MAST model's `.sitelh` file and identifies
the modal tree class at each genome position using a sliding-window approach (window = 20 bp).

```{r site_likelihood_analysis, fig.show='hold'}
for (cfg in threshold_configs) {

  k              <- cfg$best_k
  top_trees      <- cfg$top_trees
  pall_k         <- met.brewer("Klimt", n = k)
  tree_col_names <- paste0("tree_", top_trees)
  tree_cols      <- 3:(2 + k)
  window_size    <- 20

  # Load site likelihoods
  trimmed_sitelh <- read.table(
    paste0("../output/", cfg$output_dir,
           "/mast_model_", k, "/mast_model_final_", k, ".sitelh"),
    header = TRUE
  )
  # Columns: Site, LnL, then one per tree in weight-descending order
  colnames(trimmed_sitelh) <- c("site", "LnL", tree_col_names)

  # Relative likelihood: best minus second-best tree at each site
  long_sitelh <- pivot_longer(
    trimmed_sitelh,
    cols      = all_of(tree_col_names),
    values_to = "likelihood",
    names_to  = "tree"
  ) %>%
    group_by(site) %>%
    mutate(
      rellh = max(likelihood) - sort(likelihood, decreasing = TRUE)[2],
      rank  = rev(base::rank(likelihood))
    )

  trimmed_sitelh$rel_lh <- (long_sitelh %>% filter(rank == 1))$rellh

  # Most likely tree per site
  trimmed_sitelh$most_likely <-
    tree_col_names[apply(trimmed_sitelh[, tree_cols], 1, which.max)]

  # Sliding window: most common tree assignment in each window
  n_sites             <- nrow(trimmed_sitelh)
  most_likely_segment <- character(n_sites)

  for (i in seq_len(n_sites)) {
    seg_end <- min(i + window_size, n_sites)
    seg     <- trimmed_sitelh[i:seg_end, ]
    most_likely_segment[i] <- names(which.max(table(seg$most_likely)))
  }
  trimmed_sitelh$window_lh <- most_likely_segment

  # Modal class per site: mode of window assignments across all windows containing site i
  modal_class         <- character(n_sites)
  modal_rellikelihood <- numeric(n_sites)
  modal_proportion    <- numeric(n_sites)

  for (i in seq_len(n_sites)) {
    seg_start <- max(1,       i - (window_size - 1))
    seg_end   <- min(n_sites, i + (window_size - 1))
    seg       <- trimmed_sitelh[seg_start:seg_end, ]
    most_freq              <- names(which.max(table(seg$window_lh)))
    modal_class[i]         <- most_freq
    modal_rellikelihood[i] <- mean(seg$rel_lh, na.rm = TRUE)
    modal_proportion[i]    <- sum(seg$window_lh == most_freq) / nrow(seg)
  }

  trimmed_sitelh$modal_class <- modal_class
  trimmed_sitelh$modal_prob  <- modal_rellikelihood
  trimmed_sitelh$modal_prop  <- modal_proportion
  trimmed_sitelh$site        <- as.numeric(row.names(trimmed_sitelh))

  # Attach gappiness and entropy computed earlier
  trimmed_sitelh$windowentropy <- entropydf_list[[cfg$threshold]]$rollingmean
  trimmed_sitelh$gappyness     <- gapdf_list[[cfg$threshold]]$rollingmean

  # ---- Plots ----

  # Explicit limits (not just breaks) so every panel below plots the same genomic
  # coordinate at the same physical x-position — panels whose own data don't span
  # the full alignment (e.g. window_plot) would otherwise silently rescale to fit
  # just their own data range, throwing off alignment with the other panels.
  x_scale <- scale_x_continuous(
    expand = c(0, 0),
    limits = c(0, n_sites),
    breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 8000, 9000)
  )

  # 1. Modal class strip
  modal_class_plot <- ggplot(trimmed_sitelh, aes(x = site, y = 'tree', colour = modal_class)) +
    geom_point(size = 7, shape = 108) +
    labs(title = paste0("Modal class — threshold: ", cfg$threshold)) +
    theme(
      text             = element_text(size = 8),
      axis.title       = element_blank(),
      axis.text        = element_text(size = 8),
      axis.ticks.y     = element_blank(),
      axis.text.y      = element_blank(),
      axis.line.y      = element_blank(),
      legend.position  = 'top',
      legend.title     = element_text(size = 7),
      legend.text      = element_text(size = 7),
      legend.key.size  = unit(0.4, "lines"),
      plot.title       = element_text(size = 8),
      plot.margin      = margin(1, 1, 1, 1, "mm")
    ) +
    scale_colour_manual(values = pall_k) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    x_scale

  print(modal_class_plot)
  save_gg(modal_class_plot, paste0("../figures/", cfg$threshold, "/likelihood_modal_class_plot"),
          width = 180, height = 20, units = "mm", dpi = 300, bg = "white")

  # 1b. Originating window per top tree — where each tree's own ML topology was inferred
  # from, for comparison against where MAST calls it the modal (best-supported) class above.
  window_df      <- get_tree_windows(cfg$threshold)
  window_df      <- window_df[match(top_trees, window_df$tree_num), ]
  window_df$tree <- tree_col_names

  window_plot <- ggplot(window_df, aes(x = start, xend = end, y = tree, yend = tree, colour = tree)) +
    geom_segment(linewidth = 4, lineend = "butt") +
    scale_colour_manual(values = pall_k) +
    x_scale +
    theme(
      text            = element_text(size = 8),
      axis.title      = element_blank(),
      axis.text.x     = element_blank(),
      axis.ticks.x    = element_blank(),
      axis.text.y     = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'none',
      plot.margin     = margin(1, 1, 1, 1, "mm")
    )

  # 2. Mean relative likelihood coloured by modal class
  spline_prob <- as.data.frame(spline(trimmed_sitelh$site, trimmed_sitelh$modal_prob))

  mean_prob_plot <- ggplot(trimmed_sitelh) +
    geom_line(data = spline_prob, aes(x = x, y = y), alpha = 0.5, color = "grey50") +
    geom_point(aes(x = site, y = modal_prob, colour = modal_class), size = 0.25) +
    x_scale +
    scale_y_continuous(breaks = c(0, 1, 2, 3, 4)) +
    scale_colour_manual(values = pall_k) +
    theme(
      text            = element_text(size = 8),
      axis.title      = element_blank(),
      axis.text.y     = element_text(size = 8),
      axis.text.x     = element_blank(),
      axis.ticks.x    = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom',
      plot.margin     = margin(1, 1, 1, 1, "mm")
    ) +
    guides(colour = 'none')

  print(mean_prob_plot)
  save_gg(mean_prob_plot, paste0("../figures/", cfg$threshold, "/mean_likelihood_modal"),
          width = 173, height = 32, units = "mm", dpi = 400, bg = "white")

  # 3. Proportion of windows supporting modal class
  spline_prop <- as.data.frame(spline(trimmed_sitelh$site, trimmed_sitelh$modal_prop))

  proportion_plot <- ggplot(trimmed_sitelh) +
    geom_line(data = spline_prop, aes(x = x, y = y), color = "grey50", alpha = 0.5) +
    geom_point(aes(x = site, y = modal_prop, colour = modal_class), size = 0.25) +
    x_scale +
    scale_colour_manual(values = pall_k) +
    theme(
      text            = element_text(size = 8),
      axis.title      = element_blank(),
      axis.text.y     = element_text(size = 8),
      axis.text.x     = element_blank(),
      axis.ticks.x    = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom',
      plot.margin     = margin(1, 1, 1, 1, "mm")
    ) +
    guides(colour = 'none')

  print(proportion_plot)
  save_gg(proportion_plot, paste0("../figures/", cfg$threshold, "/proportion_modal"),
          width = 173, height = 32, units = "mm", dpi = 400, bg = "white")

  # 4. Entropy coloured by modal class
  entropy_modal_plot <- ggplot(trimmed_sitelh) +
    geom_line( aes(x = site, y = windowentropy), color = "grey50", alpha = 0.5) +
    geom_point(aes(x = site, y = windowentropy, colour = modal_class), size = 0.25) +
    x_scale +
    scale_colour_manual(values = pall_k) +
    theme(
      text            = element_text(size = 8),
      axis.title      = element_blank(),
      axis.text       = element_text(size = 8),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom',
      plot.margin     = margin(1, 1, 1, 1, "mm")
    ) +
    guides(colour = 'none')

  # 5. Gappiness coloured by modal class
  gappy_modal_plot <- ggplot(trimmed_sitelh) +
    geom_line( aes(x = site, y = gappyness), color = "grey50", alpha = 0.5) +
    geom_point(aes(x = site, y = gappyness, colour = modal_class), size = 0.25) +
    x_scale +
    scale_colour_manual(values = pall_k) +
    theme(
      text            = element_text(size = 8),
      axis.title      = element_blank(),
      axis.text.y     = element_text(size = 8),
      axis.text.x     = element_blank(),
      axis.ticks.x    = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom',
      plot.margin     = margin(1, 1, 1, 1, "mm")
    ) +
    guides(colour = 'none')

  # window_plot/modal_class_plot keep their previous absolute size (10.32mm/17.20mm);
  # the four lower panels are pinned to fixed heights - 18mm each, except the bottom
  # one (entropy_modal_plot), which carries the x-axis ticks/labels and needs more room.
  combined_plot <- (window_plot / modal_class_plot / mean_prob_plot / proportion_plot /
                     gappy_modal_plot / entropy_modal_plot) +
                    plot_layout(heights = unit(c(10.32, 17.2, 18, 18, 18, 22.3), "mm"))

  print(combined_plot)

  # 168mm was measured empirically (not calculated): patchwork adds inter-panel
  # spacing on top of each plot's own plot.margin, so the true rendered height of
  # the six panels above (fixed unit(mm) heights) exceeds their nominal sum + margins.
  # Rendering at a deliberately oversized canvas and reading back the actual content
  # extent gave ~167.5mm; a plain arithmetic sum consistently under-shoots and clips
  # the bottom panel(s).
  save_raster_and_svg(function() print(combined_plot),
                       paste0("../figures/", cfg$threshold, "/sitelh_plots"),
                       width = 165, height = 168,
                       units = "mm", res = 300)

  message("Done: site likelihood figures for threshold ", cfg$threshold)
}
```


### Tree weights

Two weight barplots per threshold: the initial 35-candidate-tree weights (before model
selection, from `mast_model_35`/`mast_35`), and the re-optimized weights of just the
best-k trees in the final selected model. Both come from the `Tree weights:` line IQ-TREE
writes into the `.iqtree` report.

```{r weights_helper}
extract_tree_weights <- function(iqtree_path) {
  line <- grep("^Tree weights", readLines(iqtree_path, warn = FALSE), value = TRUE)
  as.numeric(strsplit(gsub("Tree weights: |,", "", line), "\\s+")[[1]])
}
```

Initial 35-tree weights (unordered by tree index, and ranked descending):
```{r weights_init35, fig.show='hold'}
for (cfg in threshold_configs) {

  w          <- extract_tree_weights(
    paste0("../output/", cfg$output_dir, "/mast_model_35/mast_model_final_35.iqtree")
  )
  weights_df <- data.frame(tree = seq_along(w), weight = w)

  p_unordered <- ggplot(weights_df, aes(x = tree, y = weight)) +
    geom_col() +
    labs(
      title = paste0("Initial 35-tree weights — threshold: ", cfg$threshold),
      x = "Tree", y = "Weight"
    ) +
    theme_bw() +
    theme(text = element_text(size = 8), plot.title = element_text(size = 8))
  print(p_unordered)

  p_reordered <- ggplot(weights_df, aes(x = reorder(tree, -weight), y = weight)) +
    geom_col() +
    labs(
      title = paste0("Initial 35-tree weights (ranked) — threshold: ", cfg$threshold),
      x = "Tree", y = "Weight"
    ) +
    theme_bw() +
    theme(text = element_text(size = 8), plot.title = element_text(size = 8))
  print(p_reordered)

  save_pdf_and_svg(function() print(p_unordered),
                    paste0("../figures/", cfg$threshold, "/init35_weights"), width = 6, height = 4)

  save_pdf_and_svg(function() print(p_reordered),
                    paste0("../figures/", cfg$threshold, "/init35_weights_reordered"), width = 6, height = 4)

  message("Done: init35 weight plots for threshold ", cfg$threshold)
}
```

Best-k model weights, ranked descending with values labeled on the bars:
```{r weights_bestk, fig.show='hold'}
for (cfg in threshold_configs) {

  k <- cfg$best_k
  w <- extract_tree_weights(
    paste0("../output/", cfg$output_dir, "/mast_model_", k, "/mast_model_final_", k, ".iqtree")
  )

  weights_df         <- data.frame(tree = cfg$top_trees, weight = w)
  weights_df         <- weights_df[order(-weights_df$weight), ]
  weights_df$tree     <- factor(weights_df$tree, levels = weights_df$tree)

  weight_plot <- ggplot(weights_df, aes(x = tree, y = weight)) +
    geom_col(fill = "grey45", width = 0.9) +
    geom_text(
      aes(label = round(weight, digits = 3), y = 0.02),
      color = "white", size = 8 / .pt, vjust = 0.5
    ) +
    labs(
      title = paste0("Best-", k, " model weights — threshold: ", cfg$threshold),
      y = "Model weight", x = "Tree"
    ) +
    theme(
      text         = element_text(size = 8),
      axis.ticks   = element_blank(),
      axis.title.y = element_blank(),
      axis.text    = element_text(size = 8),
      axis.title.x = element_text(size = 8),
      plot.title   = element_text(size = 8)
    ) +
    guides(colour = 'none')

  print(weight_plot)

  # PNG+SVG at the display size; a second PDF at a different (publication) aspect ratio
  # is also kept below — same plot, so one SVG for it is enough.
  save_gg(weight_plot, paste0("../figures/", cfg$threshold, "/weights_bestk"),
          width = 65, height = 32, units = "mm", dpi = 400, bg = "white")

  pdf(paste0("../figures/", cfg$threshold, "/weights_bestk.pdf"), width = 5, height = 2)
  print(weight_plot)
  dev.off()

  message("Done: best-k weight plot for threshold ", cfg$threshold)
}
```


---

## Branch support and taxon contributions (per threshold)

For each threshold, the three highest-weighted trees of its best model are compared
pairwise: cophyloplots annotated with site concordance factors (SCF, from
`iqtree_scf.sh`/`scripts/iqtree_scf.sh <threshold> <tree1> <tree2> <tree3>`), and each
taxon's contribution to topological discordance between the pair.

### Branch support — site concordance factors (SCF)

Loading and preprocessing the three highest-weighted trees per threshold.

```{r scf_load}
load_scf_tree <- function(path) {
  tr   <- read.tree(path)
  vals <- do.call(rbind, strsplit(tr$node.label, "/"))
  tr$node.label <- as.numeric(vals[, 2])   # sCF
  tr$bootstrap  <- as.numeric(vals[, 1])   # bootstrap
  tr
}

root_and_trim <- function(tr) {
  tr <- root(tr, outgroup = "CPZ", resolve.root = TRUE)
  tr <- ladderize(tr)
  tr$tip.label <- sub("B.FR.83.HXBLAI_IIIB_BRU.K03455", "B.FR", tr$tip.label, fixed = TRUE)
  drop.tip(tr, "CPZ")
}

top3_by_weight <- function(cfg) {
  w <- extract_tree_weights(
    paste0("../output/", cfg$output_dir, "/mast_model_", cfg$best_k,
           "/mast_model_final_", cfg$best_k, ".iqtree")
  )
  # cfg$top_trees is the *initial* 35-tree run's ranking; the k-tree subset gets
  # re-optimized jointly, so its weights (and hence rank order) can differ. Take
  # the top 3 by the re-optimized weights, not the initial-run order.
  cfg$top_trees[order(-w)][1:3]
}

scf_trees_list <- list()
for (cfg in threshold_configs) {
  top3 <- top3_by_weight(cfg)
  trs  <- lapply(top3, function(i) {
    root_and_trim(load_scf_tree(
      paste0("../output/", cfg$output_dir, "/scf/scf_tree_", i, ".cf.tree")
    ))
  })
  names(trs) <- paste0("tree_", top3)
  scf_trees_list[[cfg$threshold]] <- trs
}
```

```{r scf_palette}
subtypes_of_interest <- c('A', 'B', 'C', 'D', 'AE', 'F1', 'cpx', 'J', 'BF')
assoc        <- cbind(subtypes_of_interest, subtypes_of_interest)
pall_cophylo <- met.brewer("Klimt", n = 9)
taxa_colors  <- setNames(pall_cophylo, subtypes_of_interest)
```

Palette reference:
```{r scf_palette_plot, fig.width=5, fig.height=2}
palldf      <- data.frame(tip = assoc[, 1], color = pall_cophylo)
palldf$tip  <- factor(palldf$tip, levels = palldf$tip)

ggplot(palldf, aes(x = tip, y = 1, fill = color)) +
  geom_point(shape = 22, size = 15) +
  scale_fill_identity() +
  theme_classic() +
  theme(
    text        = element_text(size = 8),
    axis.text.y = element_blank(), axis.title = element_blank(), axis.ticks = element_blank()
  )
```

Helper to add sCF low-support dots (red = sCF < 33%):
```{r scf_cophylo_helper}
add_scf_dots <- function(obj) {
  scf_left  <- as.numeric(obj$trees[[1]]$node.label)
  scf_right <- as.numeric(obj$trees[[2]]$node.label)
  nodelabels.cophylo(
    node = which(scf_left  < 33) + Ntip(obj$trees[[1]]),
    pch = 19, col = "red", cex = 0.5
  )
  nodelabels.cophylo(
    node = which(scf_right < 33) + Ntip(obj$trees[[2]]),
    pch = 19, col = "red", cex = 0.5, which = "right"
  )
}
```

### Cophyloplots and taxon contributions to discordance

```{r taxon_contrib_fn}
taxon.contribution.to.disc <- function(tr1, tr2) {
  taxa        <- intersect(tr1$tip.label, tr2$tip.label)
  disc.metric <- setNames(numeric(length(taxa)), taxa)
  for (i in seq_along(taxa)) {
    disc.metric[i] <- MutualClusteringInfo(drop.tip(tr1, taxa[i]), drop.tip(tr2, taxa[i]))
  }
  disc.metric
}

make_contrib_df <- function(x) {
  data.frame(subt = factor(names(x), levels = names(x)), contrib = x)
}

contrib_theme <- theme(
  text               = element_text(size = 8),
  panel.grid.major.y = element_line(color = "grey90"),
  axis.ticks         = element_blank(),
  axis.text          = element_text(size = 8),
  legend.position    = "none",
  axis.title         = element_blank(),
  plot.title         = element_text(size = 8)
)
```

```{r cophylo_and_taxon_contrib, fig.show='hold'}
for (cfg in threshold_configs) {

  top3  <- top3_by_weight(cfg)
  trs   <- scf_trees_list[[cfg$threshold]]
  pairs <- combn(top3, 2, simplify = FALSE)

  contrib_plots <- list()

  for (pr in pairs) {
    t1    <- trs[[paste0("tree_", pr[1])]]
    t2    <- trs[[paste0("tree_", pr[2])]]
    label <- paste0(pr[1], "v", pr[2])

    # Cophyloplot with SCF dots
    obj <- cophylo(t1, t2, assoc = assoc, rotate = FALSE, use.edge.length = TRUE)
    draw_cophylo <- function() {
      plot(obj, part = 0.42, link.col = pall_cophylo, link.lwd = 1.5, link.type = "curved",
           link.lty = "solid", fsize = c(0.8, 0.8), ftype = 'off', pts = FALSE)
      add_scf_dots(obj)
    }
    save_raster_and_svg(draw_cophylo,
                         paste0("../figures/", cfg$threshold, "/", label, "_cophyloplot_w_scf"),
                         width = 63, height = 54, units = "mm", res = 400)

    # Taxon contribution to discordance
    disc       <- taxon.contribution.to.disc(t1, t2)
    contrib_df <- make_contrib_df(sort(disc[subtypes_of_interest]))
    pad        <- 0.05 * diff(range(contrib_df$contrib))

    p <- ggplot(contrib_df, aes(x = subt, y = contrib, fill = subt)) +
      geom_col() + scale_fill_manual(values = taxa_colors) +
      coord_cartesian(ylim = range(contrib_df$contrib) + c(-pad, pad)) +
      contrib_theme + ggtitle(paste0("Trees ", pr[1], " vs ", pr[2]))

    print(p)
    contrib_plots[[label]] <- p

    # Individual SVG per pair (the combined PDF below is multi-page, which SVG can't hold).
    contrib_svg <- paste0("../figures/", cfg$threshold, "/", label, "_taxon_contribution.svg")
    ggsave(contrib_svg, plot = p, device = "svg", width = 57, height = 30, units = "mm", bg = "white")
    set_svg_mm(contrib_svg, 57, 30)
  }

  pdf(
    paste0("../figures/", cfg$threshold, "/Fig_taxon_contribution.pdf"),
    width = 57 / 25.4, height = 30 / 25.4, useDingbats = FALSE
  )
  for (p in contrib_plots) print(p)
  dev.off()

  message("Done: SCF cophyloplots + taxon contribution for threshold ", cfg$threshold)
}
```
