---
title: "Post processing and visualising MAST results - all thresholds"
output: html_document
date: "2025-03-25"
author: "Sofie Agerbæk"
---

```{r libraries, include = FALSE}
library(Biostrings)
library(ape)
library(seqinr)
library(tidyverse)
library(dplyr)
library(TreeDist)
library(MetBrewer)
library(phytools)
library(patchwork)
library(zoo)
```

```{r theme_setup, include = FALSE}
library(cowplot)
options(digits = 3)
options(pillar.sigfig = 3)

text_base_size <- 12
fig.witdh      <- 180  # mm
fig.height     <- 125  # mm

theme_set(theme_cowplot(font_size = 12, rel_small = 1, rel_tiny = 1, rel_large = 1))
knitr::opts_chunk$set(fig.width  = fig.witdh  / 25.4)
knitr::opts_chunk$set(fig.height = fig.height / 25.4)
knitr::opts_chunk$set(dpi = 600)

ggplot_text_size <- text_base_size / ggplot2::.pt
```

```{r palettes, include = FALSE}
pall_global <- met.brewer("Klimt", n = 6)
options(ggplot2.discrete.fill   = pall_global)
options(ggplot2.discrete.colour = pall_global)
options(ggplot2.continuous.colour = scales::seq_gradient_pal(low = "white", high = "#023F62"),
        ggplot2.continuous.fill   = scales::seq_gradient_pal(low = "white", high = "#023F62"))
```


## Threshold configuration

One entry per threshold. To add or change a threshold, edit this chunk only — everything
downstream reads from `threshold_configs`.

- `threshold`:   the gap threshold string (also used in input/output filenames)
- `output_dir`:  folder name inside `output/`
- `best_k`:      optimal number of trees by BIC minimum (verified against BIC CSV below)
- `top_trees`:   tree numbers in the best model, in descending weight order
                 (from the iterative MAST run in `iqtree_modelselect.R`)

Trees in descending weight order per threshold (all 35):
0.5 → 30 8 31 10 2 21 6 9 4 24 14 29 16 26 19 32 20 13 1 23 25 7 12 35 28 33 27 5 17 11 18 22 3 34 15
0.9 → 5 33 25 17 24 9 22 2 15 4 32 35 10 ...
0.1 → 28 22 27 13 9 8 16 23 7 26 ...

```{r threshold_config}
threshold_configs <- list(
  list(
    threshold  = "0.1",
    output_dir = "0.1",
    best_k     = 4,           # BIC minimum at n=4 (178240.07)
    top_trees  = c(28, 22, 27, 13)
  ),
  list(
    threshold  = "0.5",
    output_dir = "0.5_output",
    best_k     = 6,           # BIC minimum at n=6 (173132.59)
    top_trees  = c(30, 8, 31, 10, 2, 21)
  ),
  list(
    threshold  = "0.9",
    output_dir = "0.9",
    best_k     = 5,           # BIC minimum at n=5 (163767.40)
    top_trees  = c(5, 33, 25, 17, 24)
  )
)
```

Sanity check — verify that the hardcoded `best_k` matches the BIC minimum in each CSV:
```{r bic_sanity_check}
for (cfg in threshold_configs) {
  bic <- read.csv(paste0("../output/", cfg$output_dir, "/bic_values.csv"), header = TRUE)
  auto_k <- bic$n[which.min(bic$value)]
  if (auto_k != cfg$best_k) {
    warning(sprintf(
      "Threshold %s: hardcoded best_k = %d but BIC minimum is at n = %d. Check the config!",
      cfg$threshold, cfg$best_k, auto_k
    ))
  } else {
    message(sprintf("Threshold %s: best_k = %d confirmed by BIC data.", cfg$threshold, cfg$best_k))
  }
}
```


---

## Per-threshold analyses

### BIC vs. number of trees

```{r bic_plots, fig.show='hold'}
for (cfg in threshold_configs) {

  BIC_table <- read.csv(
    paste0("../output/", cfg$output_dir, "/bic_values.csv"),
    header = TRUE
  ) %>% mutate(BIC_transformed = value / 10000)

  bic_plot <- ggplot(BIC_table, aes(x = n, y = BIC_transformed)) +
    geom_point(colour = "black", size = 1.2) +
    geom_vline(xintercept = cfg$best_k, linetype = "dashed", colour = "grey50") +
    theme_bw() +
    labs(title = paste0("Gap threshold: ", cfg$threshold, "  |  best k = ", cfg$best_k)) +
    theme(
      axis.ticks          = element_blank(),
      panel.grid.minor    = element_blank(),
      axis.text           = element_text(size = 12, colour = "black"),
      axis.title          = element_blank()
    )

  print(bic_plot)

  ggsave(
    paste0("../figures/", cfg$threshold, "_BIC_plot.png"),
    plot = bic_plot, device = "png",
    width = 90, height = 60, units = "mm", dpi = 400, bg = "white"
  )
}
```


### Gappiness and entropy of alignments

```{r gappy_entropy_plots, fig.show='hold'}

# Store per-threshold data for re-use in the site likelihood section below
gapdf_list     <- list()
entropydf_list <- list()

site_entropy_fn <- function(M, gaps = c("-", "N", "n"), base = 2) {
  apply(M, 2, function(col) {
    col <- col[!(col %in% gaps)]
    if (!length(col)) return(NA_real_)
    p <- prop.table(table(col))
    -sum(p * log(p, base = base))
  })
}

for (cfg in threshold_configs) {

  aln_path <- paste0("../data/", cfg$threshold, "_trimmed_alns_HIV_2021_subset.fasta")
  window   <- 20

  # --- Gappiness ---
  aln     <- readDNAMultipleAlignment(aln_path)
  aln_mat <- as.matrix(aln)
  gaps    <- colSums(aln_mat == "-") / nrow(aln_mat)

  gapdf           <- data.frame(gap_prop = gaps)
  gapdf$site      <- as.numeric(row.names(gapdf))
  gapdf$rollingmean <- zoo::rollapply(gapdf$gap_prop, width = window, mean, partial = TRUE)

  # --- Entropy ---
  dna       <- read.dna(aln_path, format = "fasta")
  H_nogap   <- site_entropy_fn(as.character(dna))
  entropydf <- data.frame(H_nogap = H_nogap)
  entropydf$site        <- as.numeric(rownames(entropydf))
  entropydf$rollingmean <- zoo::rollapply(entropydf$H_nogap, width = window, mean, fill = NA)

  # Store for site likelihood section
  gapdf_list[[cfg$threshold]]     <- gapdf
  entropydf_list[[cfg$threshold]] <- entropydf

  # --- Plots ---
  gapprop_plot <- ggplot(gapdf) +
    geom_point(aes(x = site, y = gap_prop), size = 0.7, color = "grey80") +
    geom_line( aes(x = site, y = rollingmean), color = "grey50", alpha = 0.9) +
    scale_x_continuous(expand = c(0,0),
                       breaks = c(0,1000,2000,3000,4000,5000,6000,7000,8000,9000)) +
    scale_y_continuous(breaks = c(0, 0.25, 0.5)) +
    theme(
      axis.title   = element_blank(),
      axis.ticks.y = element_blank(),
      axis.line.x  = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.x  = element_blank()
    )

  entropy_plot <- ggplot(entropydf, aes(x = site)) +
    geom_point(aes(y = H_nogap), size = 0.7, alpha = 0.4, color = "grey80") +
    geom_line( aes(y = rollingmean), color = "grey50", alpha = 0.9) +
    scale_x_continuous(expand = c(0,0),
                       breaks = c(0,1000,2000,3000,4000,5000,6000,7000,8000,9000)) +
    scale_y_continuous(breaks = c(0, 1, 2)) +
    theme(
      axis.title       = element_blank(),
      axis.ticks.y     = element_blank(),
      legend.position  = 'bottom'
    ) +
    guides(colour = 'none')

  comb_plot <- gapprop_plot / entropy_plot
  print(comb_plot)

  ggsave(
    paste0("../figures/", cfg$threshold, "_gappy_entropy.png"),
    comb_plot, device = 'png',
    width = 190, height = 70, units = 'mm', dpi = 300, bg = 'white'
  )
}
```


### Tree distances (MDS of best-model trees)

```{r tree_distance_plots, fig.show='hold'}
for (cfg in threshold_configs) {

  k         <- cfg$best_k
  top_trees <- cfg$top_trees
  pall_k    <- met.brewer("Klimt", n = k)

  trees <- read.tree(
    paste0("../output/", cfg$output_dir,
           "/mast_model_", k, "/mast_model_final_", k, ".treefile")
  )

  pair_dist <- MutualClusteringInfo(trees)
  mds_trees <- cmdscale(as.dist(pair_dist))

  mds_df <- data.frame(
    x       = mds_trees[, 1],
    y       = mds_trees[, 2],
    tree_nr = as.character(top_trees)
  )

  tree_dist_plot <- ggplot(mds_df, aes(x = x, y = y, colour = tree_nr)) +
    geom_point(size = 3) +
    labs(
      title  = paste0("Tree distances — threshold: ", cfg$threshold),
      colour = NULL, y = NULL, x = NULL
    ) +
    scale_colour_manual(values = pall_k) +
    theme(axis.ticks = element_blank())

  print(tree_dist_plot)

  ggsave(
    paste0("../figures/", cfg$threshold, "_tree_distances.png"),
    plot = tree_dist_plot, device = "png",
    width = 85, height = 55, units = "mm", dpi = 400, bg = "white"
  )
}
```


### Site likelihood — modal classes

For each threshold, reads the best-k MAST model's `.sitelh` file and identifies
the modal tree class at each genome position using a sliding-window approach (window = 20 bp).

```{r site_likelihood_analysis, fig.show='hold'}
for (cfg in threshold_configs) {

  k              <- cfg$best_k
  top_trees      <- cfg$top_trees
  pall_k         <- met.brewer("Klimt", n = k)
  tree_col_names <- paste0("tree_", top_trees)
  tree_cols      <- 3:(2 + k)
  window_size    <- 20

  # Load site likelihoods
  trimmed_sitelh <- read.table(
    paste0("../output/", cfg$output_dir,
           "/mast_model_", k, "/mast_model_final_", k, ".sitelh"),
    header = TRUE
  )
  # Columns: Site, LnL, then one per tree in weight-descending order
  colnames(trimmed_sitelh) <- c("site", "LnL", tree_col_names)

  # Relative likelihood: best minus second-best tree at each site
  long_sitelh <- pivot_longer(
    trimmed_sitelh,
    cols      = all_of(tree_col_names),
    values_to = "likelihood",
    names_to  = "tree"
  ) %>%
    group_by(site) %>%
    mutate(
      rellh = max(likelihood) - sort(likelihood, decreasing = TRUE)[2],
      rank  = rev(base::rank(likelihood))
    )

  trimmed_sitelh$rel_lh <- (long_sitelh %>% filter(rank == 1))$rellh

  # Most likely tree per site
  trimmed_sitelh$most_likely <-
    tree_col_names[apply(trimmed_sitelh[, tree_cols], 1, which.max)]

  # Sliding window: most common tree assignment in each window
  n_sites             <- nrow(trimmed_sitelh)
  most_likely_segment <- character(n_sites)

  for (i in seq_len(n_sites)) {
    seg_end <- min(i + window_size, n_sites)
    seg     <- trimmed_sitelh[i:seg_end, ]
    most_likely_segment[i] <- names(which.max(table(seg$most_likely)))
  }
  trimmed_sitelh$window_lh <- most_likely_segment

  # Modal class per site: mode of window assignments across all windows containing site i
  modal_class         <- character(n_sites)
  modal_rellikelihood <- numeric(n_sites)
  modal_proportion    <- numeric(n_sites)

  for (i in seq_len(n_sites)) {
    seg_start <- max(1,       i - (window_size - 1))
    seg_end   <- min(n_sites, i + (window_size - 1))
    seg       <- trimmed_sitelh[seg_start:seg_end, ]
    most_freq              <- names(which.max(table(seg$window_lh)))
    modal_class[i]         <- most_freq
    modal_rellikelihood[i] <- mean(seg$rel_lh, na.rm = TRUE)
    modal_proportion[i]    <- sum(seg$window_lh == most_freq) / nrow(seg)
  }

  trimmed_sitelh$modal_class <- modal_class
  trimmed_sitelh$modal_prob  <- modal_rellikelihood
  trimmed_sitelh$modal_prop  <- modal_proportion
  trimmed_sitelh$site        <- as.numeric(row.names(trimmed_sitelh))

  # Attach gappiness and entropy computed earlier
  trimmed_sitelh$windowentropy <- entropydf_list[[cfg$threshold]]$rollingmean
  trimmed_sitelh$gappyness     <- gapdf_list[[cfg$threshold]]$rollingmean

  # ---- Plots ----

  x_scale <- scale_x_continuous(
    expand = c(0, 0),
    breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 8000, 9000)
  )

  # 1. Modal class strip
  modal_class_plot <- ggplot(trimmed_sitelh, aes(x = site, y = 'tree', colour = modal_class)) +
    geom_point(size = 12, shape = 108) +
    labs(title = paste0("Modal class — threshold: ", cfg$threshold)) +
    theme(
      axis.title       = element_blank(),
      axis.ticks.y     = element_blank(),
      axis.text.y      = element_blank(),
      axis.line.y      = element_blank(),
      legend.position  = 'top'
    ) +
    scale_colour_manual(values = pall_k) +
    x_scale

  print(modal_class_plot)
  ggsave(
    paste0("../figures/", cfg$threshold, "_likelihood_modal_class_plot.png"),
    modal_class_plot, device = 'png',
    width = 180, height = 20, units = 'mm', dpi = 300, bg = 'white'
  )

  # 2. Mean relative likelihood coloured by modal class
  spline_prob <- as.data.frame(spline(trimmed_sitelh$site, trimmed_sitelh$modal_prob))

  mean_prob_plot <- ggplot(trimmed_sitelh) +
    geom_line(data = spline_prob, aes(x = x, y = y), alpha = 0.5, color = "grey50") +
    geom_point(aes(x = site, y = modal_prob, colour = modal_class), size = 0.7) +
    x_scale +
    scale_y_continuous(breaks = c(0, 1, 2, 3, 4)) +
    scale_colour_manual(values = pall_k) +
    theme(
      axis.title      = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom'
    ) +
    guides(colour = 'none')

  print(mean_prob_plot)
  ggsave(
    paste0("../figures/", cfg$threshold, "_mean_likelihood_modal.png"),
    mean_prob_plot, device = 'png',
    width = 173, height = 32, units = 'mm', dpi = 400, bg = 'white'
  )

  # 3. Proportion of windows supporting modal class
  spline_prop <- as.data.frame(spline(trimmed_sitelh$site, trimmed_sitelh$modal_prop))

  proportion_plot <- ggplot(trimmed_sitelh) +
    geom_line(data = spline_prop, aes(x = x, y = y), color = "grey50", alpha = 0.5) +
    geom_point(aes(x = site, y = modal_prop, colour = modal_class), size = 0.7) +
    x_scale +
    scale_colour_manual(values = pall_k) +
    theme(
      axis.title      = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom'
    ) +
    guides(colour = 'none')

  print(proportion_plot)
  ggsave(
    paste0("../figures/", cfg$threshold, "_proportion_modal.png"),
    proportion_plot, device = 'png',
    width = 173, height = 32, units = 'mm', dpi = 400, bg = 'white'
  )

  # 4. Entropy coloured by modal class
  entropy_modal_plot <- ggplot(trimmed_sitelh) +
    geom_line( aes(x = site, y = windowentropy), color = "grey50", alpha = 0.5) +
    geom_point(aes(x = site, y = windowentropy, colour = modal_class), size = 0.7) +
    x_scale +
    scale_colour_manual(values = pall_k) +
    theme(
      axis.title      = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom'
    ) +
    guides(colour = 'none')

  # 5. Gappiness coloured by modal class
  gappy_modal_plot <- ggplot(trimmed_sitelh) +
    geom_line( aes(x = site, y = gappyness), color = "grey50", alpha = 0.5) +
    geom_point(aes(x = site, y = gappyness, colour = modal_class), size = 0.7) +
    x_scale +
    scale_colour_manual(values = pall_k) +
    theme(
      axis.title      = element_blank(),
      axis.ticks.y    = element_blank(),
      legend.position = 'bottom'
    ) +
    guides(colour = 'none')

  combined_plot <- modal_class_plot / mean_prob_plot / proportion_plot /
                   gappy_modal_plot / entropy_modal_plot

  print(combined_plot)

  png(
    paste0("../figures/", cfg$threshold, "_sitelh_plots.png"),
    height = 6, width = 8, units = "in", res = 300
  )
  print(combined_plot)
  dev.off()

  message("Done: site likelihood figures for threshold ", cfg$threshold)
}
```


---

## 0.5-threshold-specific analyses

> **Note:** Site concordance factors (SCF) have currently only been computed for the
> 0.5 threshold (see `iqtree_scf.sh`). The cophyloplots and taxon contribution plots
> below are therefore specific to the 0.5 best model (trees 10, 21, 30).
> To produce equivalent figures for 0.1 or 0.9, first update `iqtree_scf.sh` with
> the relevant alignment and tree paths and re-run it.

### Branch support — site concordance factors (SCF)

Loading and preprocessing the three highest-weighted trees from the 0.5 best model.

```{r scf_load_05}
load_scf_tree <- function(path) {
  tr   <- read.tree(path)
  vals <- do.call(rbind, strsplit(tr$node.label, "/"))
  tr$node.label <- as.numeric(vals[, 2])   # sCF
  tr$bootstrap  <- as.numeric(vals[, 1])   # bootstrap
  tr
}

tree10_scf <- load_scf_tree("../output/0.5_output/scf/scf_tree_10.cf.tree")
tree21_scf <- load_scf_tree("../output/0.5_output/scf/scf_tree_21.cf.tree")
tree30_scf <- load_scf_tree("../output/0.5_output/scf/scf_tree_30.cf.tree")

root_and_trim <- function(tr) {
  tr <- root(tr, outgroup = "CPZ", resolve.root = TRUE)
  tr <- ladderize(tr)
  tr$tip.label <- sub("B.FR.83.HXBLAI_IIIB_BRU.K03455", "B.FR", tr$tip.label, fixed = TRUE)
  drop.tip(tr, "CPZ")
}

ladd_10_scf <- root_and_trim(tree10_scf)
ladd_21_scf <- root_and_trim(tree21_scf)
ladd_30_scf <- root_and_trim(tree30_scf)
```

```{r scf_palette_05}
subtypes_of_interest <- c('A', 'B', 'C', 'D', 'AE', 'F1', 'cpx', 'J', 'BF')
assoc        <- cbind(subtypes_of_interest, subtypes_of_interest)
pall_cophylo <- met.brewer("Klimt", n = 9)
taxa_colors  <- setNames(pall_cophylo, subtypes_of_interest)
```

Palette reference:
```{r scf_palette_plot_05, fig.width=5, fig.height=2}
palldf      <- data.frame(tip = assoc[, 1], color = pall_cophylo)
palldf$tip  <- factor(palldf$tip, levels = palldf$tip)

ggplot(palldf, aes(x = tip, y = 1, fill = color)) +
  geom_point(shape = 22, size = 15) +
  scale_fill_identity() +
  theme_classic() +
  theme(axis.text.y = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())
```

Helper to add sCF low-support dots (red = sCF < 33%):
```{r scf_cophylo_helper}
add_scf_dots <- function(obj) {
  scf_left  <- as.numeric(obj$trees[[1]]$node.label)
  scf_right <- as.numeric(obj$trees[[2]]$node.label)
  nodelabels.cophylo(
    node = which(scf_left  < 33) + Ntip(obj$trees[[1]]),
    pch = 19, col = "red", cex = 0.5
  )
  nodelabels.cophylo(
    node = which(scf_right < 33) + Ntip(obj$trees[[2]]),
    pch = 19, col = "red", cex = 0.5, which = "right"
  )
}
```

Cophyloplots:
```{r cophylo_10v21_05}
png("../figures/10v21_cophyloplot_w_scf.png", height = 7, width = 8, res = 400, units = "cm")
obj <- cophylo(ladd_10_scf, ladd_21_scf, assoc = assoc, rotate = FALSE, use.edge.length = TRUE)
plot(obj, part=0.42, link.col=pall_cophylo, link.lwd=1.5, link.type="curved",
     link.lty="solid", fsize=c(0.8,0.8), ftype='off', pts=FALSE)
add_scf_dots(obj)
dev.off()
```

```{r cophylo_10v30_05}
png("../figures/10v30_cophyloplot_w_scf.png", height = 7, width = 8, res = 400, units = "cm")
obj <- cophylo(ladd_10_scf, ladd_30_scf, assoc = assoc, rotate = FALSE, use.edge.length = TRUE)
plot(obj, part=0.42, link.col=pall_cophylo, link.lwd=1.5, link.type="curved",
     link.lty="solid", fsize=c(0.8,0.8), ftype='off', pts=FALSE)
add_scf_dots(obj)
dev.off()
```

```{r cophylo_21v30_05}
png("../figures/21v30_cophyloplot_w_scf.png", height = 7, width = 8, res = 400, units = "cm")
obj <- cophylo(ladd_21_scf, ladd_30_scf, assoc = assoc, rotate = FALSE, use.edge.length = TRUE)
plot(obj, part=0.42, link.col=pall_cophylo, link.lwd=1.5, link.type="curved",
     link.lty="solid", fsize=c(0.8,0.8), ftype='off', pts=FALSE)
add_scf_dots(obj)
dev.off()
```


### Taxon contributions to discordance

```{r taxon_contrib_05}
taxon.contribution.to.disc <- function(tr1, tr2) {
  taxa        <- intersect(tr1$tip.label, tr2$tip.label)
  disc.metric <- setNames(numeric(length(taxa)), taxa)
  for (i in seq_along(taxa)) {
    disc.metric[i] <- MutualClusteringInfo(drop.tip(tr1, taxa[i]), drop.tip(tr2, taxa[i]))
  }
  disc.metric
}

make_contrib_df <- function(x) {
  data.frame(subt = factor(names(x), levels = names(x)), contrib = x)
}

disc_10_21 <- taxon.contribution.to.disc(ladd_10_scf, ladd_21_scf)
disc_10_30 <- taxon.contribution.to.disc(ladd_10_scf, ladd_30_scf)
disc_21_30 <- taxon.contribution.to.disc(ladd_21_scf, ladd_30_scf)

hiv_contrib_tab_10_21 <- make_contrib_df(sort(disc_10_21[subtypes_of_interest]))
hiv_contrib_tab_10_30 <- make_contrib_df(sort(disc_10_30[subtypes_of_interest]))
hiv_contrib_tab_21_30 <- make_contrib_df(sort(disc_21_30[subtypes_of_interest]))

contrib_theme <- theme(
  panel.grid.major.y = element_line(color = "grey90"),
  axis.ticks         = element_blank(),
  legend.position    = "none",
  axis.title         = element_blank()
)

p1 <- ggplot(hiv_contrib_tab_10_21, aes(x = subt, y = contrib, fill = subt)) +
  geom_col() + scale_fill_manual(values = taxa_colors) +
  coord_cartesian(ylim = c(8.1, 8.8)) + contrib_theme + ggtitle("Trees 10 vs 21")

p2 <- ggplot(hiv_contrib_tab_10_30, aes(x = subt, y = contrib, fill = subt)) +
  geom_col() + scale_fill_manual(values = taxa_colors) +
  coord_cartesian(ylim = c(6.8, 7.5)) + contrib_theme + ggtitle("Trees 10 vs 30")

p3 <- ggplot(hiv_contrib_tab_21_30, aes(x = subt, y = contrib, fill = subt)) +
  geom_col() + scale_fill_manual(values = taxa_colors) +
  coord_cartesian(ylim = c(7.3, 8.0)) + contrib_theme + ggtitle("Trees 21 vs 30")

print(p1); print(p2); print(p3)

pdf("../figures/Fig_taxon_contribution.pdf", width = 5, height = 1.5, useDingbats = FALSE)
print(p1); print(p2); print(p3)
dev.off()
```
