---
title: "Make ML trees"
output: html_document
date: "2025-08-27"
author: "Sofie Agerbæk"
---


Load alignment, trim data, subset data into partitions 
For each partition make a ML tree, save these as newick files.

```{r, echo=FALSE}
library(seqinr)
library(ape)
library(ggplot2)
library(purrr)
library(dplyr)
library(phytools)
```

## Load alignment
```{r}
genome_alignment <- read.alignment("../data/HIV1_CON_2021_genome_DNA_subset.fasta", format = "fasta")
alignment_matrix <- as.DNAbin(genome_alignment)
```

## Trim data - remove columns with gaps in more than X % of the sequences.  

Define trimming function
```{r}
trimCols <- function(al, prop, codon = T){
  mat <- as.character(as.matrix(al))
  ntax <- nrow(mat)
  propthres <- 1-prop
  compliantSites <- apply(mat, 2, function(x){
    x <- as.character(x)
    compl <- (length(which(x %in% c("N", "n", "?", "-", "O", "o", "X", "x"))) / ntax) < propthres
    return(compl)
  })
  
  if(codon){
    codIDs <- rep(1:(length(compliantSites)/3), each = 3)
    codsToKeep <- rep(F, length(compliantSites))
    for(i in 1:max(codIDs)){
      if(all(compliantSites[which(codIDs == i)])) codsToKeep[which(codIDs == i)] <- T
    }
    al <- al[, as.logical(codsToKeep)]
    return(al)
  } else {
    al <- al[, as.logical(compliantSites)]
    return(al)
  }
}
```

Initial run we had the threshold at 0.5, two other thresholds were also tested: 0.1 and 0.9
```{r}
threshold = 0.1 #change here for the gap 'allowance', we run 0.1, 0.5, and 0.9
trimmed_alignment <- trimCols(alignment_matrix, threshold, codon = FALSE)
```

## save trimmed alignment in fasta format
```{r}
sequences <- as.character(as.list(trimmed_alignment))
sequences <- sapply(sequences, paste, collapse = "")
headers <- names(sequences)
```

Write trimmed alignment to file
```{r}
seq_along(sequences)

sequences[1]

fasta_lines <- unlist(lapply(names(sequences), function(label) {
  seq <- paste0(sequences[[label]], collapse = "") # Join the sequence
  c(paste0(">", label), seq) # Create a header and sequence entry
}))

output_file <- "../data/0.1_trimmed_alns_HIV_2021_subset.fasta" #change output file naming when changing threshold

writeLines(fasta_lines, output_file)
```

## Make Maximum likelihood trees with IQtree 

Find IQtree 
```{r}
Sys.setenv(PATH = paste("/opt/homebrew/bin", Sys.getenv("PATH"), sep=":"))
```

Read trimmed alignment from fasta file
```{r}
genome_alignment <- read.alignment("../data/0.1_trimmed_alns_HIV_2021_subset.fasta", format = "fasta")
alignment_matrix <- as.DNAbin(genome_alignment)
```


Partitioning data into 18 segments (~500bp segment for each tree) and running ML to generate trees for each segment

```{r}
setwd("../generated/0.1/18_ML_trees")

alignment_length <- ncol(alignment_matrix)
num_sequences <- nrow(alignment_matrix)
window_size <- ceiling(alignment_length / 18) #rounding up to ensure the full alignment is contained in the 18 segments, as consequence the last segment may be a few nucleotides shorter than the rest. 

trees18 <- list()

missing_by_segment <- list()

for (start in seq(1, alignment_length, by = window_size)) {
  end <- start + window_size - 1
  print(start)
  
  if (end > alignment_length) {
    end <- alignment_length
  } #Make sure the last segment is not out of bounds
  
  #print(end)
  #current window 
  segment <- alignment_matrix[, start:end]
 
  # Because of the threshold allowing up to 90% gaps (threshold = 0.1), we can end up with alignment segments for which all positions are gaps for certain sequences. IQ-tree cannot handle these and will not produce a tree, so if any subtype is all gaps in any segment, we need to remove it from the alignment
  all_gaps <- names(which(checkDNA(segment) == window_size)) # Check if any sequences are all gaps
  seg_id <- paste0(start, "_", end)
  missing_by_segment[[seg_id]] <- all_gaps # To know afterwards which sequences were excluded from which segments
  
  #segment_filtered <- segment[!(rownames(segment) %in% all_gaps), ]
  
  #write.FASTA(segment_filtered, file = paste0("aln_segment_", start, "_to_", end, ".fasta")) 
  file_path <- paste0("aln_segment_", start, "_to_", end) #change filepath here for different thresholds

  #system(paste0("iqtree3 -s ", file_path, ".fasta -m MFP -B 1000 -nt 2 -pre ", file_path))
  
  ml_tree <- read.tree(paste0(file_path, ".treefile")) #read the trees written to file
  
  trees18[[paste0("Tree_", start, "_to_", end)]] <- ml_tree
}

multiPhylo <- do.call(c, trees18)
write.tree(multiPhylo, file = "all_18_trimmed_ML_trees.newick")

```

For the mast model downstream we need all taxa present in all trees, since we had to remove some taxa from some segments as they were all gaps and the ML tree building doesn't accept this, we have to add back those removed taxa to those trees. 
This is only relevant for the 0.1 threshold segment 1 
```{r}
tree1 <- trees18[["Tree_1_to_531"]]
 
internal_nodes <- (Ntip(tree1) + 1):(Ntip(tree1) + Nnode(tree1))

all_tiplabels <- trees18[["Tree_532_to_1062"]][["tip.label"]]
tree1_tiplabels <- trees18[["Tree_1_to_531"]][["tip.label"]]

missing_tips <- all_tiplabels[!all_tiplabels %in% tree1_tiplabels]

n_tree <- tree1
for (t in missing_tips) {
   n_tree <- bind.tip(n_tree, t, where = sample(internal_nodes, 1), edge.length = 0) 
   print(t)
  }

write.tree(n_tree, "tree1_test.newick")
trees18[["Tree_1_to_531"]] <- n_tree

multiPhylo <- do.call(c, trees18)
write.tree(multiPhylo, file = "all_18_trimmed_ML_trees.newick")
```


We shift the windows by half a window length to create 17 trees for comparison of break points
```{r}
setwd("../generated/0.1/17_ML_trees") ### Change this dir accordingly

half_window_size <- ceiling(window_size/2)
trees17 <- list()

for (start in seq(half_window_size, alignment_length - half_window_size + 1, by = window_size)) {
  end <- start + window_size - 1
  print(start)
  print(end)
  
  #current window 
  segment <- alignment_matrix[, start:end]
  
  write.FASTA(segment, file = paste0("aln_segment_", start, "_to_", end, ".fasta"))
  file_path <- paste0("aln_segment_", start, "_to_", end)
 
  system(paste0("iqtree3 -s ", file_path, ".fasta -m MFP -B 1000 -T 4 -pre ", file_path))
  
  ml_tree <- read.tree(paste0(file_path, ".treefile"))
  trees17[[paste0("Tree_", start, "_to_", end)]] <- ml_tree
}

#Write file with all trees in multiphylo object
multiPhylo <- do.call(c, trees17)
write.tree(multiPhylo, file = "17_shifted_windows_trimmed_ML_trees.newick")

```

```{r}
for (start in seq(half_window_size, alignment_length - half_window_size + 1, by = window_size)) {
  end <- start + window_size - 1
  print(start)
  print(end)
  print(end - start)
  }
```

Combine all trees into a single file
```{r}
trees35 <- c(trees18, trees17)
multiPhylo <- do.call(c, trees35)

setwd("~/Library/CloudStorage/OneDrive-RegionSjælland/Mol_phylo/revision/mast_hiv/generated")
write.tree(multiPhylo, file = "0.1/35_trimmed_ML_trees.newick") 
```

## Visualising branch lengths 

```{r}
trees_18_ML <- read.tree(file = "0.1/18_ML_trees/all_18_trimmed_ML_trees.newick")

branch_data18 <- imap_dfr(trees_18_ML, ~{
  tibble(
    tree = .y,                      # tree index
    branch_length = .x$edge.length  # branch lengths
  )
})

p18 <- ggplot(branch_data18)+
  geom_histogram(aes(x=log(branch_length)), bins = 30, colour = "grey60")+
  labs(title = "18 original ML trees")+
  theme_minimal()+
  ylim(0,250)

p18

#Sanity check to make sure no tree has too different branch lengths
ggplot(branch_data18, aes(x = factor(tree), y = log(branch_length), fill = factor(tree))) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    x = "Tree",
    y = "Log Branch length",
    fill = "Tree",
    title = "Branch length distributions per tree"
  ) +
  theme(legend.position = "none") 

```

```{r}
trees_17_ML <- read.tree(file = "0.1/17_ML_trees/17_shifted_windows_trimmed_ML_trees.newick")

branch_data17 <- imap_dfr(trees_17_ML, ~{
  tibble(
    tree = .y,                      # tree index
    branch_length = .x$edge.length  # branch lengths
  )
})

p17 <- ggplot(branch_data17)+
  geom_histogram(aes(x=log(branch_length)), bins = 30, colour = "grey60")+
  labs(title = "17 shifted ML trees")+
  theme_minimal()+
  ylim(0,250)
p17
```

```{r}
library(patchwork)
p18 + p17 
```



